---
title: "MLN maize loss prediction"
author: "Sebastian Palmas"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

#Introduction

#Analysis

## Packages
```{r, message=FALSE, warning=FALSE}
library(terra)
library(tidyverse)
```


## SPAM and agroecological zones

The basis of the analysis will be the SPAM maize production. Version 1.1.

SPAM production is measured in metric tons. This dataset has a 10 x 10 km resolution.

```{r}
SPAM2010_prod <- rast("F:/Work/SPAM/spam2010v1r1/spam2010v1r1_global_prod.geotiff/spam2010V1r1_global_P_MAIZ_A_KEN.tif")
```

### Correction of SPAM for different yeara
SPAM is a prediction of the production (and other variables) for 2010. Without any correction of this, we would compare losses using only 2010 production. To solve this issue, we correct the SPAM surfaces so that the total production matches the numbers reported in FAO for 2013, 2017 and 2018 for Maize.

We assume

* The area with maize production has not changed in all years
* All pixels change the same rate. In other words, there are no areas that have higher or lower proportional production between years.

* SPAM 2010 total production: 3,096,039 MT (sum(values(SPAM2010_prod), na.rm=TRUE))
* FAOSTAT estimate for 2010: 3,464,541 MT
* FAOSTAT estimate for 2013: 3,592,688 MT
* FAOSTAT estimate for 2017: 3,688,090 MT
* FAOSTAT estimate for 2018: 4,013,777 MT

In this code below, we scale the 2010 production using the values above
```{r}
#values to make the correction
SPAM2010_prod_sum <- 3096039 #(sum(values(SPAM2010_prod), na.rm=TRUE))
FAOSTAT2013_prod_sum <- 3592688
FAOSTAT2017_prod_sum <- 3688090
FAOSTAT2018_prod_sum <- 4013777

SPAM2013_prod <- SPAM2010_prod * FAOSTAT2013_prod_sum/SPAM2010_prod_sum
SPAM2017_prod <- SPAM2010_prod * FAOSTAT2017_prod_sum/SPAM2010_prod_sum
SPAM2018_prod <- SPAM2010_prod * FAOSTAT2018_prod_sum/SPAM2010_prod_sum

#writing the files
writeRaster(SPAM2013_prod, "../data/tif/spam2013_P_MAIZ_A_KEN.tif", overwrite=TRUE)
writeRaster(SPAM2017_prod, "../data/tif/spam2017_P_MAIZ_A_KEN.tif", overwrite=TRUE)
writeRaster(SPAM2018_prod, "../data/tif/spam2018_P_MAIZ_A_KEN.tif", overwrite=TRUE)
```


### Agroecological zones

This is the agrocological zones defined by X. We will use them to summarize the results.

```{r}
AEZ <- vect("F:/Work/Kenya/Agroecological zones/aez2b/aez.shp") %>% 
  project(crs = crs(SPAM2010_prod), col = NAME_AEZ)
```

### Physical area decicated to production

We need the area dedicated to maize in each agroecological zone. WE use the phys_area from SPAM to calculate this. It is measured in Hectares

These numbers are just a little different than those reported in the FAW paper, probably because of the differences in the SPAM 2010 version.

```{r}
SPAM2010_phys_area <- rast("F:/Work/SPAM/spam2010v1r1/spam2010v1r1_global_phys_area.geotiff/spam2010V1r1_global_A_MAIZ_A.tif")  #In Hectares

SPAM2010_phys_area_AEZ <- terra::extract(x = SPAM2010_phys_area, y = AEZ, fun = sum, na.rm=TRUE)  #adding all losses for that AEZ IN HECTARES
colnames(SPAM2010_phys_area_AEZ) <- "PHYS_AREA_MAIZ"

results <- cbind.data.frame(NAME_AEZ = AEZ$NAME_AEZ_E, SPAM2010_phys_area_AEZ, LR_perc = AEZ$LR_perc)
results
```



### Long rains, short rains

`spam2010V1r1_global_P_MAIZ_A_KEN.tif` is a measure of annual prodution of maize. Therefore we need to divide the annual production between the two seasons: long rains (MAR-MAY with harvest occurring October-November) and short rains(OCT-DEC with harvest ocurring February-March).
We divided them using a calculated Long rain percentage of the production. This proportion was calculated...

In order to use it in calculations, we create a raster using the Long rain proportions by AEZ using the template of SPAM. When we multiply this by the amount of production (SPAM), we get the total production for that season (or by 1-LR_perc in the case of short rains).

```{r}
AEZ_LRperc <- terra::rasterize(x=AEZ, y=SPAM2010_prod, field=AEZ$LR_perc,
                               filename="../data/tif/aez_LRperc.tif", overwrite=TRUE)
```


## Production loss
We calculate total yild loss by multiplying the annual production by the LR/SR percentage by the Yloss prediction (krigging).

We have data for 4 seasons of measurements: 2013 (annual), March 2017 (LR), October 2017 (SR) and March 2018 (LR) based on their planting date. Below we calculate total yield loss for each season in the data.

Because of the different seasons, we also correct by the proportion of the production that is LR, SR or annual in the case of 2013. 

```{r, warning=FALSE, message=FALSE}
#2013 data. Annual data. No need to correct for LR/SR percentage
mln_YLoss <- rast("../output/tif/mln_YLoss_2013_Krigging.tif") #percentage of yield reduction in the pixel/
mln_total_Loss <- SPAM2013_prod * (mln_YLoss/100) #/100 to convert from percentage to proportion
writeRaster(mln_total_Loss, filename = "../output/tif/mln_total_Loss_2013.tif", overwrite=TRUE)  #production loss in the season (tonnes/pixel)

#March 2017. LR season
mln_YLoss <- rast("../output/tif/mln_YLoss_march2017_Krigging.tif") #percentage of yield reduction in the pixel/
mln_total_Loss <- SPAM2017_prod * AEZ_LRperc * (mln_YLoss/100) #/100 to convert from percentage to proportion
writeRaster(mln_total_Loss, filename = "../output/tif/mln_total_Loss_march2017.tif", overwrite=TRUE)  #production loss in the season (tonnes/pixel)

#Oct 2017. SR season (1-LRperc)
mln_YLoss <- rast("../output/tif/mln_YLoss_oct2017_Krigging.tif") #percentage of yield reduction in the pixel/
mln_total_Loss <- SPAM2018_prod * (1 - AEZ_LRperc) * (mln_YLoss/100) #/100 to convert from percentage to proportion
writeRaster(mln_total_Loss, filename = "../output/tif/mln_total_Loss_oct2017.tif", overwrite=TRUE)  #production loss in the season (tonnes/pixel)

#March 2018. LR season
mln_YLoss <- rast("../output/tif/mln_YLoss_march2018_Krigging.tif") #percentage of yield reduction in the pixel/
mln_total_Loss <- SPAM2018_prod * AEZ_LRperc * (mln_YLoss/100) #/100 to convert from percentage to proportion
writeRaster(mln_total_Loss, filename = "../output/tif/mln_total_Loss_march2018.tif", overwrite=TRUE)  #production loss in the season (tonnes/pixel)
```


## Summary of results by AEZ
```{r}
#All season names to use in the loop
seasons  <- c("2013", "march2017", "oct2017", "march2018")

for (season in seasons){
  #season <- seasons[1]
  mln_perc_Loss <-  rast(paste0("../output/tif/mln_YLoss_", season, "_Krigging.tif")) #percentage of yield reduction in the pixel/
  mln_total_Loss <- rast(paste0("../output/tif/mln_total_Loss_", season, ".tif"))  #reading total loss raster
  
  #adding to the table
  mln_perc_Loss_AEZ <- terra::extract(x = mln_perc_Loss, y = AEZ, fun = mean, na.rm=TRUE)  #Mean loss
  colnames(mln_perc_Loss_AEZ) <- paste0("mln_mean_Loss_", season)  #changing name for the final results table
  
  mln_total_Loss_AEZ <- terra::extract(x = mln_total_Loss, y = AEZ, fun = sum, na.rm=TRUE)  #Production loss (tonnes)
  colnames(mln_total_Loss_AEZ) <- paste0("mln_total_Loss_", season)  #changing name for the final results table
  
  results <- cbind.data.frame(results, mln_perc_Loss_AEZ, mln_total_Loss_AEZ)
}

#exporting results
write.table(results, file = "../output/tables/MLN_loss_byAEZ.csv",sep=",", row.names = FALSE)

results
```

## 2013 and 2018 total loss surfaces

For representation purposes, we need a layer of results for the complete production in 2018 by adding the October 2017, March 2017 seasons.

We need two surfaces
* Percentage of production loss in 2018 = (mln_total_Loss_oct2017 + mln_total_Loss_march2018)/SPAM2010_prod
* Production loss (kg/ha). For 2018 = (mln_total_Loss_oct2017 + mln_total_Loss_march2018).

```{r}
mln_total_Loss_2013 <- rast("../output/tif/mln_total_Loss_2013.tif")
mln_total_Loss_oct2017 <- rast("../output/tif/mln_total_Loss_oct2017.tif")
mln_total_Loss_march2018 <- rast("../output/tif/mln_total_Loss_march2018.tif")

#estimating loss percentage of the production
mln_perc_Loss_2013 <- 100*mln_total_Loss_2013/SPAM2013_prod
mln_perc_Loss_2018 <- 100*(mln_total_Loss_oct2017 + mln_total_Loss_march2018)/SPAM2018_prod
writeRaster(mln_perc_Loss_2013, filename = "../output/tif/mln_perc_Loss_2013.tif", overwrite=TRUE)
writeRaster(mln_perc_Loss_2018, filename = "../output/tif/mln_perc_Loss_2018.tif", overwrite=TRUE)

#Total 2018 production is the sum of oct2017 and march 2018 seasons
mln_total_Loss_2018 <- mln_total_Loss_oct2017 + mln_total_Loss_march2018
writeRaster(mln_total_Loss_2018, filename = "../output/tif/mln_total_Loss_2018.tif", overwrite=TRUE)

```
## Total values to add as labels to the maps
```{r, echo=FALSE}
print(paste0("Percentage loss 2013 (%):", mean(values(mln_perc_Loss_2013), na.rm=TRUE)))
print(paste0("Percentage loss 2018 (%):", mean(values(mln_perc_Loss_2018), na.rm=TRUE)))

print(paste0("Total loss 2013 (MT):", sum(values(mln_total_Loss_2013), na.rm=TRUE)))
print(paste0("Total loss 2018 (MT):", sum(values(mln_total_Loss_2018), na.rm=TRUE)))
```

